Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-03-08 supplement.tex
comments to supplement

Low coverage (4-8x) Illumina short read data were generated for the Nama,
Gumuz, Amhara and Oromo populations as part of the African Diversity Reference
Panel (Sanger / Wellcome Trust) \citep{Gurdasani2015-qy,Pagani2015-pz} and
approved through a secondary data analysis agreement for this project. Briefly,
raw reads were aligned to GRCH37 with BWAmem, duplicates
marked with Picard MarkDuplicates, reads were realigned around indels with GATK
RealignerTargetCreater/IndelRealigner followed by BQSR with dbSNP 137.
Contamination checks were performed requiring that FREEMIX $<0.05$;
contamination checks resulted in the elimination of 22 Nama samples. We note
that the high heterozygosity in these genomes both due to inherent genetic
diversity and admixture may have violated base assumptions for this
heterozygosity check. Genomes were then variant called with GATK3.2 Unified
Genotyper \citep{DePristo2011-up} using joint calling across 2,478 individuals within
the ADRP dataset with a minimum base quality of 17. Data were merged with 1000
Genomes Phase 3 \citep{1000_Genomes_Project_Consortium2015-zq} using the union
of sites identified with bcftools isec (-n+1) \citep{Danecek2021-kc} then refined
with Variant Recalibrator with a truth sensitivity threshold of 99.5\%. HapMap
III and dbSNP 138 served as known sites while 1000 Genomes Phase 1 Omni2.5 and
Phase 1 genomic SNP served as the training set. After VR, no batch effects were
observed along PC1 and PC2 for 1000 Genomes vs. ADRP. Phasing on the combined
dataset was performed via SHAPEIT2 \citep{Delaneau2013-aw} and utilized the duoHMM
option for duos and trios. We highlight 82 Nama genomes which are newly
available (unpublished) under accession number EGAD00001006198. Among these 82
Nama samples, we down-sampled the individuals to minimize close relatives and
admixture, such that 44 Nama genomes were retained. 2nd and 3rd degrees
relatives were inferred from reconstructed pedigrees with Omni2.5 SNP array
data. Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
analysis, after partitioning ancestry into $k=6$ clusters with ADMIXTURE
\citep{Alexander2009-sw} where alternative ancestries represent European, West
African, Near Eastern, and East African gene flow.

@@

Given two biallelic loci, with alleles $A$ and $a$ and the left locus and
allels $B$ and $b$ at the right locus, the standard covariance measure of LD is
$D=p_{AB}p_{ab}-p_{Ab}p_{aB}$, where $p_{AB}$ is the frequency of $AB$
haplotypes in a population (and thus the probability of drawing an $AB$
haplotype in a random sample of that population). \citet{Hill1968-vu} solve for
the expectation of $D^2$ using a system of equations that includes $\E[Dz] =
\E[D(1-2p_A)(1-2p_B)]$ and $\E[\pi_2] = \E[p_A(1-p_A)p_B(1-p_B)]$, where $p_A$
and $p_B$ are the frequencies of $A$ and $B$ at the left and right loci,
respectively. This system also requires the expected pairwise diversity (or
expected heterozygosity, assuming random mating), denoted $\E[H] =
\E[2p_A(1-p_A)] = \E[2p_B(1-pB)]$, assuming equal mutation rates at the two
loci.

\citet{Ragsdale2019-nt} showed how to compute the analogous multi-population
set of LD statistics, and we refer readers there for details on their
definitions, computation, and interpretations. In short, we obtain expectations
of $D^2$ in each population, the cross-population product $D_iD_j$ (where $i$
and $j$ index populations), as well as those additional statistics $Dz$ and
$\pi_2$ taken over different combinations of population indexing. That is
$Dz_{i, j, k} = D_i(1-2p_{A,j})(1-2p_{B,k})$ and
$\pi_{2;i,j,k,l}=p_{A,i}(1-p_{A,j})p_{B,k}(1-2p_{B,l})$. We consider statistics
normalized by $\pi_2$ in a reference population (throughout, we use the Mende
$\pi_2$), which removes any dependence on the mutation rate. Thus, statistics
take the form $\sigma_{d;i,j}^2 = \frac{\E[D^2_{i,j}]}{\E[\pi_2]}$,
$\sigma_{Dz;i,j,k}=\frac{\E[Dz_{i,j,k}]}{\E[\pi_2]}$, and so on.
Multi-population pairwise diversity statistics $\E[H_i]=\E[2p_i(1-p_i)]$ and
$\E[H_{i,j}] = \E[p_i(1-p_j) + p_j(1-p_i)]$ were also normalized by $H$ in the
Mende, so that all pairwise diversity measures are relative to the reference
population.

@@

Based on the geographical locations of present-day populations, we tentatively
labeled ancestral branches using a parsimony in migration, referring to South,
East, and West African branches (S/E/W AFR,
Figures~\ref{fig:1},~\ref{fig:supp-possible-models}). We do not know the
geographical location of these ancestral populations (nor even if they
correspond to populations with a well-defined geographic range), and these
labels should be considered as tentative. However, we found it useful to name
branches in reference to where in Africa their descendants are found.
Parameters were converted to physical units
(years, effective population sizes, etc)
assuming a human generation time of 29 years.

@@

Our models also included recent events to account for known migrations,
admixtures, and growths and declines in effective population sizes. Many of
these parameters were fixed based on previous historical, genetic, or
anthropological research, namely
\begin{itemize}
    \item East African pastoralist to South African admixture: 2 ka \cite{}
    \item European to South African admixture: 10 generations ago \cite{}
    \item Mende population expansion and Gumuz population decline: 5 ka \cite{}
    \item South African population decline following colonial admixture: 9 generations ago \cite{}
\end{itemize}
While the dates of these events were fixed, the sizes and proportions were
allowed to vary in the fits. The total number of parameters that were
ultimately inferred were 21-31, depending on the complexity
of the model.

@@

Here we focus on four parameterizations of early human history: (1) a “single
origin” with recent expansion, (2) primary and second stem populations with
“continuous migration”, (3) two stem populations that do not exchange migrants
before merging to form regional African populations, referred to as a “merger
without stem migration”, and (4) two stem populations that are allowed to
exchange migrants before merging to form regional African populations, referred
to as a “merger with stem migration”. Given the best fit parameters for each
model
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
we used \msprime version 1.0.4 \citep{Kelleher2016-lw,Baumdicker2021-mj} to
simulate sampled individual genomes to match the data used in inference (see
simulation details below). Using these simulations, we compared additional
statistics and inferences between simulated and real data that were not used in
the original optimization. These comparisons serve as independent validations
of our inferred demographic models by assessing how well those models fit
aspects of the data that have been used in other studies as evidence of
population structure or archaic hominin admixture from an unidentified hominin species
within Africa.

@@

Prior studies primarily relied on the mutation clock in order to date events
such as population divergence, and uncertainty about the mutation rate
translates in uncertainty in timing estimates (e.g., \citet{Moorjani2016-qj}).
Recent whole genome sequencing studies from non-pathological pedigrees have
typically estimated mutation rates to be $1.2-1.3\times10^{-8}$ per site per
generation \citep{Sasani2019-zp,Tian2019-zv}. A rate of $1.3\times10^{-8}$ also
approximates the rate predicted by a mutational model from Gnomad
\citep{Karczewski2020-le} in the $\sim$650Mb of intergenic regions selected for
demographic inference in this study.
\texttt{moments-LD} differs from other approaches by
relying on a recombination clock, which has been argued to be more robust to
estimation than a mutational clock \citep{Moorjani2016-ur}.

From our recombination-rate-calibrated demographic models, we can estimate the
mutation rates needed to fit patterns of pairwise diversity. With our best-fit
model (merger-with-stem-migration,
Table~\ref{tab:supp-merger-with-stem-migration}), a mutation rate of
$1.33\times10^{-8}$ recovers the observed diversity within and between
populations, closely matching the average mutation rate estimated from the
Gnomad mutation model. The other inferred demographic models
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-without-stem-migration})
would require mutation rates of $1.75-1.90\times10^{-8}$ to match pairwise
diversity.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.65\textwidth]{figures/supp-possible-models.png}
    \caption{
        \textbf{Proposed models of early human history in Africa.}
    }
    \label{fig:supp-possible-models}
\end{figure}

>>>>

Low coverage (4-8x) Illumina short read data were generated for the Nama,
Gumuz, Amhara and Oromo populations as part of the African Diversity Reference
Panel (Sanger / Wellcome Trust) \citep{Gurdasani2015-qy,Pagani2015-pz} and
approved through a secondary data analysis agreement for this project. Briefly,
raw reads were aligned to GRCH37 with BWAmem, duplicates
marked with Picard MarkDuplicates, reads were realigned around indels with GATK
RealignerTargetCreater/IndelRealigner followed by BQSR with dbSNP 137.
Contamination checks were performed requiring that FREEMIX $<0.05$;
contamination checks resulted in the elimination of 22 Nama samples. We note
that the high heterozygosity in these genomes both due to inherent genetic
diversity and admixture may have violated base assumptions for this
heterozygosity check. Genomes were then variant called with GATK3.2 Unified
Genotyper \citep{DePristo2011-up} using joint calling across 2,478 individuals within
the ADRP dataset with a minimum base quality of 17. Data were merged with 1000
Genomes Phase 3 \citep{1000_Genomes_Project_Consortium2015-zq} using the union
of sites identified with bcftools isec (-n+1) \citep{Danecek2021-kc} then refined
with Variant Recalibrator with a truth sensitivity threshold of 99.5\%. HapMap
III and dbSNP 138 served as known sites while 1000 Genomes Phase 1 Omni2.5 and
Phase 1 genomic SNP served as the training set. After VR, no batch effects were
observed along PC1 and PC2 for 1000 Genomes vs. ADRP. Phasing on the combined
dataset was performed via SHAPEIT2 \citep{Delaneau2013-aw} and utilized the duoHMM
option for duos and trios. We highlight 82 Nama genomes which are newly
available (unpublished) under accession number EGAD00001006198.
\comment{This makes is sound like these weren't generated for this project.}
Among these 82
Nama samples, we down-sampled the individuals to minimize close relatives and
admixture, such that 44 Nama genomes were retained. 2nd and 3rd degrees
relatives were inferred from reconstructed pedigrees with Omni2.5 SNP array
data. Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
analysis, after partitioning ancestry into $k=6$ clusters with \texttt{ADMIXTURE}
\comment{Justification for k=6}
\citep{Alexander2009-sw} where alternative ancestries represent European, West
African, Near Eastern, and East African gene flow.

@@

Given two biallelic loci, with alleles $A$ and $a$ and the left locus and
alleles $B$ and $b$ at the right locus, the standard covariance measure of LD is
$D=p_{AB}p_{ab}-p_{Ab}p_{aB}$, where $p_{AB}$ is the frequency of $AB$
haplotypes in a population (and thus the probability of drawing an $AB$
haplotype in a random sample of that population). \citet{Hill1968-vu} solve for
the expectation of $D^2$ using a system of equations that includes $\E[Dz] =
\E[D(1-2p_A)(1-2p_B)]$ and $\E[\pi_2] = \E[p_A(1-p_A)p_B(1-p_B)]$, where $p_A$
and $p_B$ are the frequencies of $A$ and $B$ at the left and right loci,
respectively. This system also requires the expected pairwise diversity (or
expected heterozygosity, assuming random mating), denoted $\E[H] =
\E[2p_A(1-p_A)] = \E[2p_B(1-pB)]$, assuming equal mutation rates at the two
loci.

\citet{Ragsdale2019-nt} showed how to compute the analogous multi-population
set of LD statistics, and we refer readers there for details on their
definitions, computation, and interpretations. In short, we obtain expectations
of $D^2$ in each population, the cross-population product $D_iD_j$ (where $i$
and $j$ index populations), as well as those additional statistics $Dz$ and
$\pi_2$ taken over different combinations of population indexing. That is
$Dz_{i, j, k} = D_i(1-2p_{A,j})(1-2p_{B,k})$ and
$\pi_{2;i,j,k,l}=p_{A,i}(1-p_{A,j})p_{B,k}(1-2p_{B,l})$. We consider statistics
normalized by $\pi_2$ in a reference population (throughout, we use the Mende
\comment{Why use the Mende to normalize?}
$\pi_2$), which removes any dependence on the mutation rate. Thus, statistics
take the form $\sigma_{d;i,j}^2 = \frac{\E[D^2_{i,j}]}{\E[\pi_2]}$,
$\sigma_{Dz;i,j,k}=\frac{\E[Dz_{i,j,k}]}{\E[\pi_2]}$, and so on.
Multi-population pairwise diversity statistics $\E[H_i]=\E[2p_i(1-p_i)]$ and
$\E[H_{i,j}] = \E[p_i(1-p_j) + p_j(1-p_i)]$ were also normalized by $H$ in the
Mende, so that all pairwise diversity measures are relative to the reference
population.

@@

Based on the geographical locations of present-day populations, we tentatively
labeled ancestral branches using a parsimony in migration, referring to South,
East, and West African branches
\comment{would also be interesting to organize by ethnolinguistic group...}
(S/E/W AFR, Figures~\ref{fig:1},~\ref{fig:supp-possible-models}). We do not know the
geographical location of these ancestral populations (nor even if they
correspond to populations with a well-defined geographic range), and these
labels should be considered as tentative. However, we found it useful to name
branches in reference to where in Africa their descendants are found.
Parameters were converted to physical units
(years, effective population sizes, etc)
assuming a human generation time of 29 years.

@@

Our models also included recent events to account for known migrations,
admixtures, and growths and declines in effective population sizes. Many of
these parameters were fixed based on previous historical, genetic, or
anthropological research, namely
\begin{itemize}
    \item East African pastoralist to South African admixture: 2 ka \cite{}
    \item European to South African admixture: 10 generations ago \cite{}
    \item Mende population expansion and Gumuz population decline: 5 ka \cite{}
    \item South African population decline following colonial admixture: 9 generations ago \cite{}
\end{itemize}
\comment{This could be organized with a bit more structure. E.g. by region, or by timepoint}
While the dates of these events were fixed, the sizes and proportions were
allowed to vary in the fits. The total number of parameters that were
ultimately inferred were 21-31, depending on the complexity
of the model.

@@

Here we focus on four parameterizations of early human history: (1) a “single
origin” with recent expansion, (2) primary and second stem populations with
“continuous migration”, (3) two stem populations that do not exchange migrants
before merging to form regional African populations, referred to as a “merger
without stem migration”, and (4) two stem populations that are allowed to
exchange migrants before merging to form regional African populations, referred
to as a “merger with stem migration”. Given the best fit parameters for each
model
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
we used \msprime version 1.0.4 \citep{Kelleher2016-lw,Baumdicker2021-mj} to
simulate sampled individual genomes to match the data used in inference (see
simulation details below). Using these simulations, we compared additional
statistics and inferences between simulated and real data that were not used in
the original optimization.
\comment{Describe in more detail}
These comparisons serve as independent validations
of our inferred demographic models by assessing how well those models fit
aspects of the data that have been used in other studies as evidence of
population structure or archaic hominin admixture from an unidentified hominin species
within Africa.

@@

Prior studies primarily relied on the mutation clock in order to date events
such as population divergence, and uncertainty about the mutation rate
translates in uncertainty in timing estimates (e.g., \citet{Moorjani2016-qj}).
Recent whole genome sequencing studies from non-pathological pedigrees have
typically estimated mutation rates to be $1.2-1.3\times10^{-8}$ per site per
generation \citep{Sasani2019-zp,Tian2019-zv}. A rate of $1.3\times10^{-8}$ also
approximates the rate predicted by a mutational model from gnomAD
\citep{Karczewski2020-le} in the $\sim$650Mb of intergenic regions selected for
demographic inference in this study.
\texttt{moments-LD} differs from other approaches by
relying on a recombination clock, which has been argued to be more robust to
estimation than a mutational clock \citep{Moorjani2016-ur}.

From our recombination-rate-calibrated demographic models, we can estimate the
mutation rates needed to fit patterns of pairwise diversity. With our best-fit
model (merger-with-stem-migration,
Table~\ref{tab:supp-merger-with-stem-migration}), a mutation rate of
$1.33\times10^{-8}$ recovers the observed diversity within and between
populations, closely matching the average mutation rate estimated from the
gnomAD mutation model. The other inferred demographic models
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-without-stem-migration})
would require mutation rates of $1.75-1.90\times10^{-8}$ to match pairwise
diversity.

@@

\clearpage
\comment{30 is a lot of supp figures... can these be condensed to be a bit less daunting to readers?}

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.65\textwidth]{figures/supp-possible-models.png}
    \caption{
        \textbf{Proposed models of early human history in Africa.}
        \comment{Lovely figure, may even want to move to main text}
    }
    \label{fig:supp-possible-models}
\end{figure}
